Imputation of Incomplete Multilevel Data with \proglang{R}

### 2023-12-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{ulem}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
MCAR & Missing Completely At Random, where the probability to be missing
is equal \\
& across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed \\
& information \\
MNAR & Missing Not At Random (MNAR), where the probability to be
missing \\
& depends on unrecorded information, making the missingness
non-ignorable \\
& \citep{rubi76, meng94}. \\
& \\
\end{longtable}

@@

Prior to implementing multiple imputation, it is critical to assess the
robustness of the estimation model when confronted with missing data.
\cite{Alisson_2002} highlighted scenarios in which specific Maximum
Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
methods. Notably, when the response variable is the sole incomplete
variable, mixed models demonstrate robustness to missing data under the
Missing at Random (MAR) assumption and with a correct
variance-covariance specification.\cite{Molenberghs_2007}. Furthermore,
depending on the proportion of missingness (usually is \textless5\%),
opting for a simpler complete case analysis might be sufficient.

@@

Various strategies can be employed in the imputation process to account
for clustering, including the inclusion of a cluster indicator variable,
conducting a separate imputation process for each cluster, or employing
a simultaneous imputation method that considers clustering (Stata:
Clustering and MI Impute) TODO: Replace ref.

The choice of strategy depends primarily on the assumptions made in the
main analysis and the constraints imposed by the analyzed
data.Concerning data restrictions, if there are few clusters with many
observations per cluster, the cluster indicator or imputation separated
by groups may be appropriate \cite{Graham_2009}. Conversely, an
imputation model that simultaneously imputes all clusters using a
hierarchical model can be employed when there are more clusters or fewer
observations per cluster \cite{Allison_2002}.

In a hierarchical imputation model, correlations between observations
within clusters are modeled by a random effect, allowing estimation even
with a limited number of observations per cluster. Multiple imputation
models based on hierarchical models have been proposed, each with
different assumptions.

Before choosing a particular strategy, it is crucial to verify if the
assumptions of the imputation model align with those of the main
analysis. For example, using the cluster indicator strategy may
introduce bias estimates when the model is based on a hierarchical
structure \cite{Taaljard_2008}. Even if an imputation strategy congruent
with the main model is preferred, it is essential to assess its
appropriateness for the data, considering that less complex imputation
strategies may lead to unbiased estimates in certain scenarios
\textbackslash cite(Bailey\_2020).

@@

The imputation model must be congenial with the main model
\cite{Meng_1994}. Congeniality issues arise when the imputation model
and the main model make different assumptions, often due to the omission
of a polynomial or interaction term or the use of transformed variables.

The imputation model can include additional terms than the main model
terms that do not lead to uncongeniality. For instance, it is advisable
to include the outcome variable in the imputation model for prediction
variables \cite{Moons}. In cases where the outcome is time-to-event, the
Nelson-Aalen estimate of the time to the event should be included as a
covariate in the imputation model {[}REF{]}. Furthermore, the inclusion
of auxiliary variables, even if not part of the main model, can be
associated with the probability of missingness, enhancing the likelihood
of satisfying the Missing at Random (MAR) assumption and improving
estimation efficiency \cite{Hardt_2012}.

The specification of imputation models is done on a variable basis,
either using a prediction matrix where the type of prediction variable
should be specified accordingly (see table) or through a list of
formulas in the formula parameter. The latter option proves useful in
formulating imputation models with polynomial terms or interactions
compared to the prediction matrix specification, which requires the
inclusion of additional terms as Just Another Variable (JAV).

However, even when considering an interaction term, depending on the
adapted individual imputation model, it has been suggested, for
instance, for treatment interaction effects, to conduct separated
imputation by treatment group \cite{Zhang_2023}. Additionally, there are
imputation models based on random forest or deep learning that can
handle interaction and non-linear terms that do not require the explicit
inclusion of the non-linear terms.

@@

The research objective of the \texttt{popmis} dataset is to predict the
pupils' popularity based on their gender and the experience of the
teacher. The analysis model corresponding to this dataset is multilevel
regression with random intercepts, random slopes and a cross-level
interaction. The outcome variable is \texttt{popular}, which is
predicted from the unit-level variable \texttt{sex} and the
cluster-level variable \texttt{texp}:

@@

The analysis model for this dataset is a prediction model with
\texttt{mort} as the outcome. In this tutorial we'll estimate the
adjusted prognostic effect of \texttt{ct} on mortality outcomes. The
estimand is the adjusted odds ratio for \texttt{ct}, after including
\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
into the analysis model:

@@

\begin{CodeChunk}
\begin{CodeInput}
R> #data("data_heckman", package = "micemd")
R> #dat <- data_heckman
\end{CodeInput}
\end{CodeChunk}

@@

We are also contemplating the utilisation of the ???pmm??? option, as
the values imputed using a fully parametric method may be implausibly
low for some patients.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mnar <- meth_mar
R> meth_mnar["Weight"]<- "3l.2stage.heckman"
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{additional}{%
\section{Additional}\label{additional}}

The imputation of these data is based on the
\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
Heckman Github repo}

Visualize missing data pattern:

The matrix only shows the predictors for the main model, not the
selection model.

TODO: explain exclusion restriction.

ORDER:

\begin{itemize}
\item
  summary
\item
  congeniality, then in hierarchical models
\item
  look whether we can fit cong. back in the main body
\item
  alt. methods
\item
  conclusion: mice is really easy!
\item
  Additional levels of clustering
\item
  More complex data types: timeseries and polynomial relationship in the
  clustering.
\item
  FIML vs MI
\end{itemize}

An alternative approach to missing data is to use Full Information
Maximum Likelihood (FIML). This method does not require the imputation
of any missing values. Whereas MI consists of imputation, analyses and
pooling steps, FIML analyses the data in a single step. When the
assumptions are met the two approaches should produce equivalent
results. {[}REF{]} As FIML requires specialised software, not all
analyses can be performed with standard software. {[}REF{]}

\begin{itemize}
\tightlist
\item
  Survival / TTE, this could be put in the paragraph on congeniality
\end{itemize}

When the outcome is time-to-event, the Nelson-Aalen estimate of the time
to event should be included as a covariate in the imputation model
{[}REF{]}

ORDER:

\begin{itemize}
\item
  summary
\item
  congeniality, then in hierarchical models
\item
  look whether we can fit cong. back in the main body
\item
  alt. methods
\item
  conclusion: mice is really easy!
\item
  Additional levels of clustering
\item
  More complex data types: timeseries and polynomial relationship in the
  clustering.
\item
  FIML vs MI
\end{itemize}

An alternative approach to missing data is to use Full Information
Maximum Likelihood (FIML). This method does not require the imputation
of any missing values. Whereas MI consists of imputation, analyses and
pooling steps, FIML analyses the data in a single step. When the
assumptions are met the two approaches should produce equivalent
results. {[}REF{]} As FIML requires specialised software, not all
analyses can be performed with standard software. {[}REF{]}

\begin{itemize}
\tightlist
\item
  Survival / TTE, this could be put in the paragraph on congeniality
\end{itemize}

When the outcome is time-to-event, the Nelson-Aalen estimate of the time
to event should be included as a covariate in the imputation model
{[}REF{]}

In hierarchical datasets, clustering is a concern because the
homoscedasticity in the error terms cannot be assumed across clusters
and the relationship among variables may vary at different hierarchical
levels. When multiple imputation is used to deal with missing data, as
the imputation and analysis process is performed separately, it is
necessary that imputation model being congenial with the main analysis
model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
structure also imputation model should do it (Audigier, 2021). Not
including clustering into the imputation process may lead to effect
estimates with smaller standard errors and inflated type I error.

There are different strategies that can be adopted in the imputation
process that account for clustering: inclusion of cluster indicator
variable, performing a separate imputation process for each cluster, or
performing a simultaneous imputation process by using an imputation
method that accounts for clustering.(Stata:
\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
TODO: replace ref.

The selection of each strategy depends mainly on the assumptions in the
main analysis and also on the restriction of the analyzed data.

Regarding the restrictions imposed by the data, for instance, the use of
cluster indicator variables is restricted in datasets where there are
not many clusters and many observations per cluster (Graham, 2009). The
last restriction is also required when imputations are performed on each
cluster separately. When this restriction cannot be achieved, one can
use an imputation model that simultaneously imputes all clusters using a
hierarchical model (Allison 2002).

Under this hierarchical imputation model, observations within clusters
are correlated and this correlation is modeled by a random effect so the
hierarchical model can be estimated even when there are few observations
per cluster. However, this strategy is best suited for balanced data
(Grund, 2017) and when random effects model is appropriated, i.e.~the
number of clusters is adequate. (Austin,2018).

Here it is important to evaluate the assumptions imposed by the main
model, for instance by using the cluster indicator strategy may lead to
bias estimates when the model is based on a hierarchical model
(Taaljard,2008). Even when an imputation strategy congenial with the
main model is preferred, it is important to consider whether it is
appropriate for the data as a less complex imputation strategies may
also lead to unbiased estimates in certain scenarios(Bailey 2020). For
instance, in causal effect analysis, separately imputation may lead to
smaller bias when the size of the smaller exposure cluster is large,
compared with an imputation model that includes exposure-confounder
interactions. (Zhang,2023).

@@

@@

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

Table 3: Notation

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Formula \pkg{lme4}}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\texttt{y\ \textasciitilde{}\ x1\ +\ (1\ \textbar{}\ g1)} & Fixed
\texttt{x1} predictor with random intercept \\
& varying among \texttt{g1} \\
\texttt{y\textasciitilde{}x1*x2+\ (1\textbar{}\ g1)} & Interactions of
\texttt{x1} and \texttt{x2} only in fixed effect \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x2\textbar{}\ g1)} & Interactions
of \texttt{x1} and \texttt{x2} only in fixed effect \\
& with slope of \texttt{x2} randomly varying among \texttt{g1} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1*x2\textbar{}\ g1)} &
variance-covariance matrix estimated only with \\
& the variance terms of intercept, slope of \texttt{x1}, \\
& slope of \texttt{x2} and interaction \texttt{x1*x2} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1\ \textbar{}\ g1)+\ (x2\textbar{}\ g1)}
& variance-covariance matrix estimated separately, \\
& i.e, one for intercept and \texttt{x1} and another for \\
& intercept and \texttt{x2} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with correlated random intercept and \\
& random slope of \texttt{x} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ \textbar{}\ g1)\ +\ (0\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with uncorrelated random intercept \\
& and random slope of \texttt{x1} \\
\texttt{y\ \textasciitilde{}\ (1\ \textbar{}\ g1)\ +\ (1\ \textbar{}\ g2)}
& Random intercept varying among \texttt{g1} and among
\texttt{g2\ \ \ \textbar{}\ \textbar{}}y \textasciitilde{} (1 \\
& \\
\end{longtable}

\bibliography{../References/multilevelmice.bib}

>>>>

\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{ulem}
\usepackage{xcolor}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
MCAR & Missing Completely At Random, where the probability to be missing
is equal \\
& across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed \\
& information \\
MNAR & Missing Not At Random (MNAR), where the probability to be
missing \\
& depends on unrecorded information, making the missingness
non-ignorable \\
& \citep{rubi76, meng1994}. \\
& \\
\end{longtable}

@@

Before diving into imputation, make sure your estimation model can
handle missing data. Sometimes, simpler methods like complete case
analysis might be suffice, especially if your missingness is low
(usually \textless5\%). There are scenarios in which specific Maximum
Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
methods, for instance when the response variable is the sole incomplete
variable, mixed models demonstrate robustness to missing data under the
Missing at Random (MAR) assumption and with a correct
variance-covariance specification.\cite{Molenberghs_2007}.

@@

When it comes to handling clusters during the imputation process, you've
got a few options. You can use a cluster indicator variable, run
separate imputation for each cluster, or go for a simultaneous
imputation method that takes clustering into account \cite{eddings}.

Which strategy you choose depends on the assumptions in your main
analysis and the limitations of your data. If your analysis do not use a
hierarchical model (like a descriptive approach) and you have a small
number of clusters with lots of observations in each, using a cluster
indicator or separate imputation might be the way to go
\cite{graham2009}. Conversely, if you have more clusters or fewer
observations per cluster, you might want to try a simultaneous
hierarchical imputation model \cite{Allison_2002}.

In a hierarchical imputation model, random effects model the
correlations between observations within clusters, making it possible to
estimate even with a small number of observations per cluster. There are
various proposed multiple imputation models based on hierarchical
models, each with its own set of assumptions \cite{audigier}.

If your analysis uses a hierarchical model, make sure the assumptions of
your imputation model match up. For example, using a cluster indicator
approach may lead to bias estimates if your model is based on a
hierarchical structure \cite{taljaard2008,speidel2018}. Even if you
prefer an imputation strategy that aligns with your main model, check if
it suits your data; sometimes simpler strategies can give unbiased
estimates in certain scenarios \cite{bailey2020}.

@@

The imputation model must be congenial with the main model
\cite{meng1994}. Congeniality issues arise when the imputation model and
the main model make different assumptions, often due to the omission of
a polynomial or interaction term or the use of transformed variables.

The imputation model can incorporate additional terms compared to the
main model without causing compatibility issues. For instance, it's
recommended to include the outcome variable in the imputation model for
prediction variables \cite{moons2006a}. In cases where the outcome is
time-to-event, the Nelson-Aalen estimate of the time to the event should
be added as a covariate in the imputation model {[}REF{]}. Additionally,
including auxiliary variables, even if not part of the main model, can
be linked to the probability of missingness, improving the likelihood of
meeting the Missing at Random (MAR) assumption and enhancing estimation
efficiency \cite{hardt2012a}.

Imputation models are specified on a variable basis, either using a
prediction matrix in the pred parameter or through a list of formulas in
the formula parameter. In the prediction matrix option, the type of each
predictor variable is specified for each incomplete variable (see
table).

\begin{align}
y_{ij} =& (\beta_0 + b_{0_i})+ \beta_1x_{1_{ij}}+\beta_2x_{2_{i.}} \nonumber \\
+& (\beta_3 +  b_{3_i})x_{3_{ij}} \nonumber \\
+& \beta_4x_{4_{ij}} + \beta_{m4}\overline{x_{4_{i.}}}\nonumber \\
+& (\beta_5 +  b_{5_i})x_{5_{ij}} +\beta_{m5}\overline{x_{5_{i.}}}\nonumber \\
\end{align}

\begin{tabular}{ll}
     \toprule
Type & Definition \\
  \midrule
-2  & Cluster variable, in this case the one defined by j index  \\
 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
 2  & Random variable, e.g., $x_{3_{ij}}$\\
 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
-3. & Random variable only included on selection model (Heckman model)\\
-4. & Random variable only included on main model (Heckman model)\\
      \bottomrule
\end{tabular}

\begin{itemize}
\tightlist
\item
  3 and 4 type have been advised to used as the inclusion of the means
  of the cluster is beneficial on FCS \cite{mistler2017}
\end{itemize}

Recipes have been proposed for imputing incomplete level-1 and level-2
variables for hierarchical models \cite{buuren2018a} (\(\S\) 7.10),
which can be convenient to follow when dealing with numerous variables
(interaction terms at different levels) and models with many random
effects that are prone to convergency problems due to overspecification
in the imputation model. These rules were designed to ensure
compatibility among the conditionally specified imputation models and
congeniality between the imputation and the main model. However, they
may not be applicable to all 2l imputation methods; for instance, the
2l.2stage imputation methods of the micemd package only allow the
inclusion of random predictor variables (2).

On the other hand, the formula option is useful in specifying complex
imputation models with polynomial terms or interactions and compared
with the prediction matrix method do not requires the inclusion of
additional terms as Just Another Variable (JAV)
\cite{buuren2018a}(\(\S\) 6.4).

For some interaction terms, it has been suggested, for instance, for
treatment interaction effects, to conduct separate imputation by
treatment group \cite{zhang2023}. Additionally, it can be used
imputation models based on random forest or deep learning that can
handle interaction and non-linear terms without requiring the explicit
specification of an imputation model.

@@

The research objective of the \texttt{popmis} dataset is to predict the
pupils' popularity based on their gender and the experience of the
teacher. The analysis model corresponding to this dataset is multilevel
regression with only random intercepts. The outcome variable is
\texttt{popular}, which is predicted from the level-1 (student) variable
\texttt{sex} and the level-2 (school) variable \texttt{texp}. Given the
\(j\)-th student belonging to the \(i\)-th school, the main model can be
formulated as:

\[ popular_{ij} = (\beta_0 + b_i) + \beta_1sex_{ij}+ \beta_2texp_{ij}+\epsilon_{ij}\]
where \(\epsilon_{ij}\) corresponds to the error term.

@@

In this dataset all the variables are level-1 (patient), except by the
level-2 (study) type variable. The analysis model for this dataset is a
prediction model with \texttt{mort} as the outcome. In this tutorial
we'll estimate the adjusted prognostic effect of \texttt{ct} on
mortality outcomes. The estimand is the adjusted odds ratio for
\texttt{ct}, after including \texttt{type}, \texttt{age}
\texttt{motor\_score} and \texttt{pupil}. Therefore the main model for
the \(i\)-th patient from the \(j\)-th study can be described as:

\[ mort_{ij} = (\beta_0 + b_i) + \beta_1type_{.j}+ \beta_2age_{ij}+ \beta_3motorscore_{ij} + \beta_4pupil_{ij} + \beta_5ct_{ij}+\epsilon_{ij}\]
where \(\epsilon_{ij}\) corresponds to the error term.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("Obesity", package = "micemd")
\end{CodeInput}
\end{CodeChunk}

@@

We are also contemplating the utilisation of the predictive mean
matching (pmm) option, as the values imputed using a fully parametric
method may be implausibly low for some patients.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mnar <- meth_mar
R> meth_mnar["Weight"]<- "2l.2stage.heckman"
\end{CodeInput}
\end{CodeChunk}

@@

@@

This paper is dedicated to exploring the imputation process for
incomplete datasets, with a primary focus on utilizing a hierarchical
model for analysis. Initially, users are encouraged to consider the main
analysis within the context of the incomplete dataset, along with
insights provided by domain experts, to gain a better understanding of
variable relationships. Employing clear data visualization is
instrumental in comprehending the missing data patterns, establishing a
missing mechanism, and aiding in the selection of suitable imputation
methods.

The ``Mice'' and ``mice''-based R packages offer a range of imputation
methods tailored for hierarchical data, easily adaptable to the
dataset's structure. Before proceeding with the analysis of the imputed
dataset, it is essential to assess the convergence of the imputation
method. This evaluation can reveal issues such as circular problems, the
need for additional iterations, or challenges associated with the chosen
imputation method.

@@

\renewcommand\refname{References}
\bibliography{../References/multilevelmice.bib}
